---
title: "Explore malaria pfpr scam fits"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
execute:
  echo: true
  message: false
  warning: false
editor: source
---

## Setup

```{r setup}
rm(list = ls())

require(glue)
require(data.table)
require(mgcv)
require(scam)
require(ggplot2)
```

## Pick a run

Defaults to the most recent dated output dir under the run root. Set `run_date_override` to e.g. `"20260511"` to look at a specific run.

```{r run-paths}
output_root       <- "/mnt/team/idd/pub/forecast-mbp/03-modeling_data/malaria/scam_fits/lsae_1285"
run_date_override <- NULL   # set to "20260511" or "20260512_v2" etc., or leave NULL

if (is.null(run_date_override)) {
  dirs <- list.dirs(output_root, recursive = FALSE)
  dirs <- dirs[grepl("^[0-9]{8}", basename(dirs))]
  stopifnot("No dated run dirs found under output_root" = length(dirs) > 0)
  output_dir <- tail(sort(dirs), 1)
} else {
  output_dir <- file.path(output_root, run_date_override)
}
stopifnot(dir.exists(output_dir))
output_dir
```

## Load all per-fit summaries

Concatenates every `summary_*.rds` in the run dir. Each is a tiny 1-row data.table written by the rocket, so this is fast even for 210 fits.

```{r load-summaries}
summary_files <- list.files(output_dir, pattern = "^summary_[0-9]+\\.csv$",
                            full.names = TRUE)
stopifnot("No summary_*.csv files in this run dir" = length(summary_files) > 0)

stats_dt <- rbindlist(lapply(summary_files, fread),
                      use.names = TRUE, fill = TRUE)
setorder(stats_dt, task_id)
message(glue("Loaded {nrow(stats_dt)} summary rows from {output_dir}"))
nrow(stats_dt)
```

## Run health

```{r health}
n_total_expected <- 210
n_present        <- nrow(stats_dt)
n_errored        <- sum(stats_dt$error, na.rm = TRUE)
n_succeeded      <- sum(!stats_dt$error, na.rm = TRUE)
n_missing        <- n_total_expected - n_present

data.table(
  metric = c("expected (param_map rows)", "summaries on disk",
             "  errored fits", "  successful fits",
             "missing (no summary written)"),
  count  = c(n_total_expected, n_present, n_errored, n_succeeded, n_missing)
)
```

### Missing task_ids

```{r missing}
present_ids <- sort(unique(stats_dt$task_id))
missing_ids <- setdiff(seq_len(n_total_expected), present_ids)
length(missing_ids)
head(missing_ids, 60)
```

### Errored fits

```{r errored}
if (any(stats_dt$error)) {
  stats_dt[error == TRUE, .(task_id, error_msg, formula_text)]
} else {
  message("No errored fits.")
}
```

## AIC leaderboard

```{r leaderboard}
leaderboard <- stats_dt[error == FALSE][order(aic)]
leaderboard[, formula_short := substr(formula_text, 1, 120)]
head(leaderboard[, .(task_id, aic, bic, dev_expl, r_sq, n_coef, edf_sum,
                     edf_smooth, n_smooths, elapsed_sec, formula_short)], 20)
```

## Distributions

### AIC

```{r aic-hist, fig.width=8, fig.height=4}
ggplot(stats_dt[error == FALSE], aes(x = aic)) +
  geom_histogram(bins = 40, fill = "steelblue", colour = "white") +
  geom_vline(xintercept = min(stats_dt$aic, na.rm = TRUE),
             colour = "red", linetype = "dashed") +
  labs(title = "AIC across all fits (red = best)", x = "AIC", y = "Count") +
  theme_minimal()
```

### Deviance explained

```{r devexpl-hist, fig.width=8, fig.height=4}
ggplot(stats_dt[error == FALSE], aes(x = dev_expl)) +
  geom_histogram(bins = 40, fill = "darkgreen", colour = "white") +
  labs(title = "Deviance explained", x = "Deviance explained", y = "Count") +
  theme_minimal()
```

### Elapsed seconds

```{r elapsed-hist, fig.width=8, fig.height=4}
ggplot(stats_dt, aes(x = elapsed_sec)) +
  geom_histogram(bins = 40, fill = "darkorange", colour = "white") +
  labs(title = "Per-fit elapsed seconds", x = "Seconds", y = "Count") +
  theme_minimal()
```

### Smoothing parameters

`max_sp` near `+Inf` (or very large) usually means a smooth was driven toward zero/linear by the penalty. Wide spread across fits is a hint that the penalty surface is doing real work.

```{r sp-scatter, fig.width=8, fig.height=4}
ggplot(stats_dt[error == FALSE & is.finite(max_sp)],
       aes(x = log10(max_sp), y = aic)) +
  geom_point(alpha = 0.6) +
  labs(title = "AIC vs log10(max smoothing parameter)",
       x = "log10(max_sp)", y = "AIC") +
  theme_minimal()
```

## Drill into specific fits

Set `inspect_task_ids` to a vector of task_ids — the chunks below load each one's full `fit_<NNN>.rds`, print summary via `mgcv:::summary.gam` (works where `summary.scam` chokes), and report the path to its plot PDF. Defaults to the top-1 by AIC; replace with whichever you want to look at.

```{r inspect-ids}
inspect_task_ids <- head(leaderboard$task_id, 1)
inspect_task_ids
```

```{r inspect-load}
load_one <- function(tid) {
  fp <- file.path(output_dir, sprintf("fit_%03d.rds", tid))
  readRDS(fp)
}
inspect_objs <- setNames(lapply(inspect_task_ids, load_one),
                         paste0("task_", inspect_task_ids))
names(inspect_objs)
```

```{r inspect-meta}
# Compact side-by-side metadata for the inspected fits.
rbindlist(lapply(inspect_objs, function(o) {
  data.table(
    task_id      = o$task_id,
    formula_text = o$formula_text,
    elapsed_sec  = o$elapsed_sec,
    error        = o$error,
    plot_pdf     = file.path(output_dir, sprintf("plot_%03d.pdf", o$task_id))
  )
}))
```

```{r inspect-summaries, results='asis'}
for (nm in names(inspect_objs)) {
  o <- inspect_objs[[nm]]
  cat(glue("\n\n### {nm}  (task_id={o$task_id})\n\n"))
  cat(glue("**Formula:** `{o$formula_text}`\n\n"))
  if (isTRUE(o$error)) {
    cat(glue("**Fit errored:** {o$error_msg}\n"))
    next
  }
  cat("```\n")
  print(mgcv:::summary.gam(o$fit))
  cat("```\n")
}
```

### Smooth term plots (inline)

If you'd rather look at the curves inside the notebook instead of the PDF, this regenerates them via `mgcv:::plot.gam`.

```{r inspect-plots, fig.width=10, fig.height=4}
for (nm in names(inspect_objs)) {
  o <- inspect_objs[[nm]]
  if (isTRUE(o$error)) next
  n_smooths <- length(o$fit$smooth)
  if (n_smooths == 0) next
  op <- par(mfrow = c(1, min(n_smooths, 4)),
            oma = c(0, 0, 3, 0))
  mgcv:::plot.gam(o$fit, pages = 1, shade = TRUE, residuals = FALSE)
  title(main = glue("{nm}: {substr(o$formula_text, 1, 100)}"),
        outer = TRUE, cex.main = 0.9)
  par(op)
}
```

## Spec structure of top-N

What `(var, form)` combinations dominate the top of the leaderboard?

```{r top-specs}
top_n     <- 20
top_files <- file.path(output_dir,
                       sprintf("fit_%03d.rds",
                               head(leaderboard$task_id, top_n)))
top_specs <- rbindlist(lapply(top_files, function(fp) {
  o <- readRDS(fp)
  if (length(o$spec) == 0) {
    return(data.table(task_id = o$task_id, var = NA_character_, form = NA_character_))
  }
  data.table(task_id = o$task_id, var = names(o$spec), form = unlist(o$spec))
}), use.names = TRUE, fill = TRUE)

top_specs[, .N, by = .(var, form)][order(-N)]
```

## Pairwise compare

```{r compare-ids}
task_a <- leaderboard$task_id[1]
task_b <- leaderboard$task_id[2]
```

```{r compare}
sa <- stats_dt[task_id == task_a]
sb <- stats_dt[task_id == task_b]
list(
  task_a    = task_a,
  task_b    = task_b,
  aic_a     = sa$aic,
  aic_b     = sb$aic,
  delta_aic = sb$aic - sa$aic,
  formula_a = sa$formula_text,
  formula_b = sb$formula_text
)
```

## Refit: IS + IS-no-FE + 5-fold country-holdout (no FE)

Given a vector of task_ids, `refit_all_7()` produces all 7 model objects per spec: the full in-sample fit with country FE, the in-sample fit without FE, and 5 country-holdout fold fits without FE. Returns a named list of lists.

```{r refit-setup}
require(arrow)

# -- Data loading (mirrors rocket) --
parquet_path      <- "/mnt/team/idd/pub/forecast-mbp/03-modeling_data/malaria/past_inputs_nc/lsae_1285/current/malaria_past_inputs.parquet"
suit_variant_pick <- "mordecai_0_0"

past_data <- as.data.frame(arrow::read_parquet(parquet_path))
past_data$A0_af <- as.factor(past_data$A0_location_id)

nan_toss <- function(df, var) {
  to_toss <- which(is.na(df[var]))
  if (length(to_toss)) df[-to_toss, ] else df
}
past_data <- nan_toss(past_data, "malaria_pfpr")
past_data <- nan_toss(past_data, "gdppc_mean")
past_data <- nan_toss(past_data, "mal_DAH_total_per_capita")

suit_col <- paste0("malaria_suitability_", suit_variant_pick)
past_data$malaria_suit_fraction     <- past_data[[suit_col]] / 365
past_data$malaria_suit_fraction     <- pmin(pmax(past_data$malaria_suit_fraction, 0.001), 0.999)
past_data$logit_malaria_suitability <- log(past_data$malaria_suit_fraction / (1 - past_data$malaria_suit_fraction))
past_data$do30_fraction <- past_data$days_over_30C / 365
past_data$do30_fraction <- pmin(pmax(past_data$do30_fraction, 0.001), 0.999)
past_data$logit_do30    <- log(past_data$do30_fraction / (1 - past_data$do30_fraction))

log_covs <- c("mal_DAH_total_per_capita", "gdppc_mean", "ldipc_mean", "med_consumppc")
for (cov in log_covs) {
  past_data[[paste0("log_", cov)]] <- log(past_data[[cov]])
}

# -- Formula builders (mirrors rocket) --
K_DEFAULT <- 6
build_term <- function(var, form) {
  if (form == "linear") return(var)
  sprintf("s(%s, k = %d, bs = '%s')", var, K_DEFAULT, form)
}
build_formula <- function(spec, response) {
  rhs <- mapply(build_term, names(spec), unlist(spec), USE.NAMES = FALSE)
  reformulate(rhs, response = response)
}

# -- Country folds (greedy bin-packing, same seed as rocket) --
N_FOLDS <- 5L
CV_SEED <- 42L

set.seed(CV_SEED)
country_ids <- unique(past_data$A0_location_id)
country_n   <- table(past_data$A0_location_id)[as.character(country_ids)]
ord         <- order(country_n, decreasing = TRUE)
country_ids <- country_ids[ord]
country_n   <- country_n[ord]

fold_sizes   <- rep(0L, N_FOLDS)
country_fold <- integer(length(country_ids))
for (i in seq_along(country_ids)) {
  smallest        <- which.min(fold_sizes)
  country_fold[i] <- smallest
  fold_sizes[smallest] <- fold_sizes[smallest] + country_n[i]
}
names(country_fold) <- as.character(country_ids)
folds <- country_fold[as.character(past_data$A0_location_id)]

message(glue("Data: {nrow(past_data)} rows, {length(country_ids)} countries"))
message(glue("Fold sizes: {paste(fold_sizes, collapse=', ')}"))
```

```{r}
dim(past_data)
hist(past_data$logit_malaria_pfpr)
```

```{r refit-function}
#' Refit all 7 models for a given task_id.
#'
#' @param tid Integer task_id
#' @return Named list: is_fit (with FE), is_no_fe_fit (without FE),
#'   cv_fits (list of 5 fold fits without FE), spec, fml, fml_no_fe
refit_7 <- function(tid) {
  spec       <- specs[[param_map[tid, spec_index]]]
  fml        <- build_formula(spec, "logit_malaria_pfpr")
  spec_no_fe <- spec[names(spec) != "A0_af"]
  fml_no_fe  <- build_formula(spec_no_fe, "logit_malaria_pfpr")

  message(glue("task_id={tid}: fitting IS (with FE) ..."))
  is_fit <- tryCatch(
    scam(fml, data = past_data, optimizer = "efs", control = list(maxit = 300)),
    error = function(e) e
  )

  message(glue("task_id={tid}: fitting IS (no FE) ..."))
  is_no_fe_fit <- tryCatch(
    scam(fml_no_fe, data = past_data, optimizer = "efs", control = list(maxit = 300)),
    error = function(e) e
  )

  message(glue("task_id={tid}: fitting 5-fold CV (no FE) ..."))
  cv_fits <- vector("list", N_FOLDS)
  for (k in seq_len(N_FOLDS)) {
    train_idx <- which(folds != k)
    cv_fits[[k]] <- tryCatch(
      scam(fml_no_fe, data = past_data[train_idx, ], optimizer = "efs",
           control = list(maxit = 300)),
      error = function(e) e
    )
    status <- if (inherits(cv_fits[[k]], "error")) "FAILED" else "OK"
    message(glue("  Fold {k}: {status}"))
  }
  names(cv_fits) <- paste0("fold_", seq_len(N_FOLDS))

  list(
    task_id      = tid,
    spec         = spec,
    fml          = fml,
    fml_no_fe    = fml_no_fe,
    is_fit       = is_fit,
    is_no_fe_fit = is_no_fe_fit,
    cv_fits      = cv_fits
  )
}

#' Refit all 7 models for a vector of task_ids.
#'
#' @param task_ids Integer vector
#' @return Named list (keyed by "task_<tid>") of refit_7() results
refit_all_7 <- function(task_ids) {
  results <- lapply(task_ids, refit_7)
  names(results) <- paste0("task_", task_ids)
  results
}
```

```{r refit-example, eval=FALSE}
# Example: refit the top 3 by OOS R²
# refit_results <- refit_all_7(head(leaderboard$task_id, 3))
# refit_results$task_42$is_fit          # full model with country FE
# refit_results$task_42$is_no_fe_fit    # model without country FE
# refit_results$task_42$cv_fits$fold_1  # fold-1 fit (no FE)
```

\`\`\`